A fully labelled proof system for intuitionistic modal logics

### 2019-10-03 caleidoscope/introduction.tex
small changes and comments everywhere

Since their introduction in the 1980s by Gabbay~\cite{gabbay:96},
\emph{labelled proof calculi} have been widely used by proof theorist
to give sound, complete and cut-free deductive systems to a broad range
of logics. Unlike so-called \emph{internal calculi}, like
hyper-sequents~\cite{avron:elc96} or nested
sequents~\cite{kashima:sl94,brunnler:aml09,poggiolesi:tmp09}, 2-sequents~\cite{masini:apal92}, or linear nested sequents~\cite{lellmann?},
labelled calculi have the advantage of being more uniform and being able to 
accommode more logics.

Standard labelled sequent calculi attach to every formla $\fm A$ a label
$\lb x$, witten as $\labels xA$, and additionally admit so called
\emph{relational atoms} of the form $\accs xy$ where $\rel$ is a
binary relation symbol. These calculi work best for logics with
standard Kripke semantics, as in this case the relation $\rel$ is used
to encode the accessibility relation in the Kripke models, and the
frame conditions corresponding to the desired logic can be directly
encoded as inference rules. Prominent examples are classical modal
logics and intuitionistic propositional logic, where, e.g., the frame
condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$), can be straightforwardly translated into the
inference rule
\begin{equation}
  \label{eq:Rtrans}
  \vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
\end{equation}
where $\B$ stands for a set of relational atoms, and $\Gamma$ and
$\Delta$ for multi-sets of labelled formulas.

However,
in this paper we are concerned with \emph{intuitionistic modal
  logics}, whose Kripke semantics is based on \emph{birelational}
frames, i.e., they have two binary relations instead of one: one
relation $R$ that corresponds to the accessibility relation in
ordinary Kripke frames for modal logics, and a relation $\le$ that
corresponds to the future relation in ordinary Kripke frames for
intuitionistic logic. Consequently, standard labelled systems for
these logics have certain shortcomings:
\begin{enumerate}
\item The transitivity rule in~\eqref{eq:Rtrans} corresponds to both
  versions of the $\fourax$-axiom
  \begin{equation}
    \label{eq:4ax}
    \fm{\BOX A\IMP \BOX\BOX A}
    \qquand
    \fm{\DIA\DIA A\IMP\DIA A}
    \quadcm
  \end{equation}
  as they are equivalent in classical
  modal logic. However, in intuitionistic modal logic they are not
  equivalent, and even though the logic $\ISfour$, i.e., the
  intuitionistic version of the modal logic $\Sfour$, contains both
  axioms, they can be added independently to the logic $\IK$ (the
  intuitionistic version of $\K$). However, the proof theory of these
  new logics has not been studied before, as no existing labelled (or
  label-free) proof system can handle them, even though the
  corresponding frame conditions
  \begin{equation}
    \label{eq:4frame}
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
    \quand
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
    \;,
  \end{equation}
  respectively, have already been studied
  in~\cite{plotkin:stirling:86}.
  \item The correspondence between the syntax and the semantics is not
  as clean as one would expect. As only the $R$-relation (and not the
  $\le$-relation) of the frame is visible in an ordinary labelled
  sequent, we only have that a sequent $\Gamma$ is provable if and
  only if is is valid in all \emph{graph-consistent}\footnote{This
    means that every layer in the model can be lifted to any future of
    any world in that layer. See~\cite{simpson:phd} and~\cite{mar:str:tableaux17} for a formal definition and discussion.} for  models, as
  already observed in by Simpson in his PhD thesis~\cite{simpson:phd}
  and considered as an inelegant solution (see also \cite{mar:str:tableaux17}).
\end{enumerate}
In order to address these two issues we propose here to extend the
notion of labelled sequent by also allowing relational atoms of the
form $\futs xy$. Consequently, we can easily translate the frame
conditions in~\eqref{eq:4frame} into inference rules:
\begin{equation}
  \label{eq:4rules}
  \vlinf{\rn{4^{\BOX}}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
     \B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
  \qquand
  \vlinf{\rn{4^{\DIA}}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
\end{equation}
This allows us to define cut-free deductive systems for a wide range
of logics that could not be treated before.  Furthermore, the relation
between syntax and semantics is as one would expect: A sequent is
provable in our system if and only if it is valid in all models.

Besides that, there is another pleasant observation to make about our
system. Ordinary labelled sequent systems for intuitionistic modal
logic are single-conclusion~\cite{simpson:phd}. The same is true for
the corresponding nested sequent
systems~\cite{str:fossacs13,marin:str:aiml}. It is possible to express
Maehara style multiple-conclusion systems in nested
sequents~\cite{kuz:str:maehara}, and therefore also in ordinary
labelled sequents. However, also in these systems there are rules
($\rlabrn\IMP$ and $\rlabrn\BOX$) that force a single-conclusion
premise, even though this is not the case in labelled
systems~\cite{negri:jpl2005} or nested sequents~\cite{fitting:83} for
intuitionistic logic. Our system eliminates this
discrepancy. Consequently, every rule in our system is invertible, as
we never delete information.

@@

\todo{maybe paragraph on related work? cite Maffezioli, Naibo and Negri~\cite{maffezioli:etal:synthese13}.}

\section*{Introduction(old version, to be deleted)}
\todo{We should work on the introduction}

One possible-world semantics was established as a solid base to define modal logics, the idea of incorporating these notions into the proof theory of modal logics emerged. Fitch seems to have been the first one to formalise it, directly including symbols representing worlds into the language of his proofs in natural deduction \cite{Fitch}.

\emph{Labelled deduction} has been proposed by Gabbay~\cite{Gabbay} in the 80s as a unifying framework throughout proof theory in order to provide proof
systems for a wide range of logics. 
%
For modal logics it can take
the form of labelled natural deduction and labelled sequent systems as
used, for example, by Simpson~\cite{Simpson}, Vigan\`o~\cite{Vigano}, and
Negri~\cite{Negri}. 
%
These formalisms make explicit use not only of
labels, but also of relational atoms referring to the accessibility relation of a Kripke frame.
%
In this short note we propose a system that represents both the \emph{accessibility relation} (for modal
logics) and the \emph{preorder relation} (for intuitionistic
logic), using the full power of the bi-relational semantics for
intuitionistic modal logics,
and developing fully the idea that Maffezioli, Naibo and Negri employed in~\cite{Maffezioli}.

\todo{Simpson does not have internal cut-elimination}

\sonia{
  \begin{itemize}
  \item internal vs externl calculi
  \item extesion to axioms are truly intuitionistic (not just the
    classical versions as in Simpson style)
  \item connection with semantics compared to graph homomorphism
    restiction in Simpson and ``indexed nested systems''
  \end{itemize}
}

\lutz{things that should be said explicitely somewhere (maybe intro, maybe somewhere else):
  \begin{itemize}
  \item what we do:
    \begin{itemize}
    \item first systems (nested and labelled) for intuitionistic modal
      logic where ``single conclusion'', i.e., only one formula on the
      right.
    \item Then ``multiple conclusion'' (Meahara-style) nested sequent.
    \item Still rules that forced single formula on the right (white implication and white box)
    \item by allowing both relations of the model in the proof system
      we go to genuinely multiple conclusion: never ever a formula on
      the right has to be deleted during proof search.
    \end{itemize}
  \item why we do it: 
    \begin{itemize}
    \item getting things right: labelled systems for classical modal
      logics, or plain propositional intuiionistic logic have one
      binary relation, which is exactly the relation of the standard
      Kripke semantics. The standard Kripke semantics for
      intuitionistic modal logic is birelational, with a strong
      connectiion between the two binary relations. A clean labelled
      system for these logics should incorporate both relations.
    \item We get new logics.  In the standard nested sequent
      or labelled sequent system for intuitionistic modal logic we
      always have both conjuncts we we add Scott-Lemmon/Geach axioms;
      but here we can have them separtely.
    \end{itemize}
  \item how we do it:
    \begin{itemize}
    \item We start from the Maehara nested sequent system (which can
      be translated straightforwardly into a labelled system with one
      relation $R$.
    \item the rules for $\AND$ and $\OR$ are trivial and as expected
      they can be presented such that they are invertible on the left
      and the right, they do not concern the relations $R$ and $\le$. 
    \item So should be the rules for $\IMP$-left and $\BOX$-left and
      identity, but there is a subtlety; see below...
    \item the rules for $\DIA$ are also straightforward, as they
      concern only $R$ but not $\le$.
    \item The rules $\BOX$-right and $\IMP$-right are the ones that
      force single conclusion in the Maehara system. This means that
      here they create a future, using the relation $\le$.
    \item In order to make everything work together, this forces us to
      have reflexivity and transitivity of $\le$, explicely as rules
      in the system, and have rules for the F1 and F2 conditions
      making the relations $R$ and $\le$ go together.
    \item As a consequence we also need a rule realizing monotonicity
      for the formulas on the right.
    \item However, we designed the system such that monotonicity is
      admisible by encorporating it into the rules for identity,
      $\IMP$-left and $\BOX$-left.
    \end{itemize}
  \end{itemize}
}
%
%For sequent systems for intuitionistic logics there is always a choice
%to be made: make the system \emph{single conclusion} following
%Gentzen~\cite{Gentzen} or \emph{multiple conclusion} following
%Maehara~\cite{Maehara}. In our work we choose the multiple conclusion
%variant because of the closer correspondence to the semantics. In that
%respect, our system is closer to~\cite{kuz:str} than
%to~\cite{Simpson} and~\cite{mar:str}.

>>>>

Since their introduction in the 1980s by Gabbay~\cite{gabbay:96},
\emph{labelled proof calculi} have been widely used by proof theorist
to give sound, complete and cut-free deductive systems to a broad range
of logics. Unlike so-called \emph{internal calculi}, like
hypersequents~\cite{avron:elc96}, nested
sequents~\cite{kashima:sl94,brunnler:aml09,poggiolesi:tmp09}, 2-sequents~\cite{masini:apal92}, or linear nested sequents~\cite{lellmann?},
labelled calculi have the advantage of being more uniform and being able to 
accommode a larger class of logics.

Standard labelled sequent calculi attach to every formla $\fm A$ a label
$\lb x$, witten as $\labels xA$, and additionally use
\emph{relational atoms} of the form $\accs xy$ where $\rel$ is a
binary relation symbol. These calculi work best for logics with
standard Kripke semantics, as in this case the relation $\rel$ is used
to encode the accessibility relation in the Kripke models, and the
frame conditions corresponding to the desired logic can be directly
encoded as inference rules. Prominent examples are classical modal
logics and intuitionistic propositional logic, where, e.g., the frame
condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$), can be straightforwardly translated into the
inference rule
\begin{equation}
  \label{eq:Rtrans}
  \vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
\end{equation}
where $\B$ stands for a set of relational atoms, and $\Gamma$ and
$\Delta$ for multi-sets of labelled formulas.

However,
in this paper we are concerned with \emph{intuitionistic modal
  logics}, whose Kripke semantics is based on \emph{birelational}
frames, i.e., they have two binary relations instead of one: one
relation $R$ that corresponds to the accessibility relation in 
Kripke frames for modal logics, and a relation $\le$ that
corresponds to the preorder relation in Kripke frames for
intuitionistic logic. Consequently, standard labelled systems for
these logics have certain shortcomings:
\begin{enumerate}
\item The transitivity rule in~\eqref{eq:Rtrans} corresponds to both
  versions of the $\fourax$-axiom
  \begin{equation}
    \label{eq:4ax}
    \fm{\BOX A\IMP \BOX\BOX A}
    \qquand
    \fm{\DIA\DIA A\IMP\DIA A}
    \quadcm
  \end{equation}
  as they are equivalent in classical
  modal logic. However, in intuitionistic modal logic they are not
  equivalent, and even though the logic $\IKfour$, i.e., the
  intuitionistic version of the modal logic $\Kfour$, contains both
  axioms, they can also be added independently to the logic $\IK$ (an
  intuitionistic version of $\K$). However, the proof theory of these
  new logics has not been studied before, as no existing labelled (or
  label-free) proof system can handle them, even though the
  corresponding frame conditions
  \begin{equation}
    \label{eq:4frame}
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
    \quand
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
    \;,
  \end{equation}
  \todo{add graphical representation of the conditions}
  respectively, have already been studied
  in~\cite{plotkin:stirling:86}.
  \sonia{the general condition is correct but not the application to 4}
  \item The correspondence between the syntax and the semantics is not
  as clean as one would expect. As only the $R$-relation (and not the
  $\le$-relation) of the frame is visible in an ordinary labelled
  sequent, we only have that a sequent $\Gamma$ is provable if and
  only if is is valid in all \emph{graph-consistent}\footnote{This
    means that every layer in the model can be lifted to any future of
    any world in that layer. See~\cite{simpson:phd} and~\cite{mar:str:tableaux17} for a formal definition and discussion.}   
  models, 
  \sonia{this is only true for logics outside of the path axioms.}
  \todo{check that Simpson's system is sound and complete for all logics defined by path axioms. Is it that the axiomatisation is also sound and complete for every models and not only graph-consistent ones?}
  as already observed in by Simpson in his PhD thesis~\cite{simpson:phd}
  and considered as an inelegant solution (see also \cite{mar:str:tableaux17}).
\end{enumerate}
In order to address these two concerns we propose here to enrich usual 
labelled sequent by allowing both, relational atoms of the
form $\futs xy$ and of the form $\accs xy$. Consequently, we can easily translate the frame
conditions in~\eqref{eq:4frame} into inference rules:
\begin{equation}
  \label{eq:4rules}
  \vlinf{\rn{4^{\BOX}}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
     \B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
  \qquand
  \vlinf{\rn{4^{\DIA}}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
\end{equation}
This allows us to define cut-free deductive systems for a wide range
of logics that could not be treated before.  Furthermore, the relation
between syntax and semantics is as one would expect: A sequent is
provable in our system if and only if it is valid in all models.

Besides that, there is another pleasant observation to make about our
system. Ordinary labelled sequent systems for intuitionistic modal
logic are single-conclusion~\cite{simpson:phd}. 
\sonia{No! the system for epistemic logic by Maffezioli, Naibo and Negri~\cite{maffezioli:etal:synthese13} is multi-conclusion}
%
The same is true for the corresponding nested sequent
systems~\cite{str:fossacs13,marin:str:aiml}. It is possible to express
Maehara style multiple-conclusion systems in nested
sequents~\cite{kuz:str:maehara}, and therefore also in ordinary
labelled sequents. However, also in these systems there are rules
($\rlabrn\IMP$ and $\rlabrn\BOX$) that force a single-conclusion
premise, even though this is not the case in labelled
systems~\cite{negri:jpl2005} or nested sequents~\cite{fitting:83} for
intuitionistic logic. Our system eliminates this
discrepancy. Consequently, every rule in our system is invertible, as
we never delete information.

@@

\todo{maybe paragraph on related work?}

%\section*{Introduction(old version, to be deleted)}
%\todo{We should work on the introduction}
%
%One possible-world semantics was established as a solid base to define modal logics, the idea of incorporating these notions into the proof theory of modal logics emerged. Fitch seems to have been the first one to formalise it, directly including symbols representing worlds into the language of his proofs in natural deduction \cite{Fitch}.
%
%\emph{Labelled deduction} has been proposed by Gabbay~\cite{Gabbay} in the 80s as a unifying framework throughout proof theory in order to provide proof
%systems for a wide range of logics. 
%%
%For modal logics it can take
%the form of labelled natural deduction and labelled sequent systems as
%used, for example, by Simpson~\cite{Simpson}, Vigan\`o~\cite{Vigano}, and
%Negri~\cite{Negri}. 
%%
%These formalisms make explicit use not only of
%labels, but also of relational atoms referring to the accessibility relation of a Kripke frame.
%%
%In this short note we propose a system that represents both the \emph{accessibility relation} (for modal
%logics) and the \emph{preorder relation} (for intuitionistic
%logic), using the full power of the bi-relational semantics for
%intuitionistic modal logics,
%and developing fully the idea that Maffezioli, Naibo and Negri employed in~\cite{Maffezioli}. 
%
%\todo{Simpson does not have internal cut-elimination}
%
%\sonia{
%  \begin{itemize}
%  \item internal vs externl calculi
%  \item extesion to axioms are truly intuitionistic (not just the
%    classical versions as in Simpson style)
%  \item connection with semantics compared to graph homomorphism
%    restiction in Simpson and ``indexed nested systems''
%  \end{itemize}
%}
%  
%\lutz{things that should be said explicitely somewhere (maybe intro, maybe somewhere else):
%  \begin{itemize}
%  \item what we do:
%    \begin{itemize}
%    \item first systems (nested and labelled) for intuitionistic modal
%      logic where ``single conclusion'', i.e., only one formula on the
%      right.
%    \item Then ``multiple conclusion'' (Meahara-style) nested sequent.
%    \item Still rules that forced single formula on the right (white implication and white box)
%    \item by allowing both relations of the model in the proof system
%      we go to genuinely multiple conclusion: never ever a formula on
%      the right has to be deleted during proof search.
%    \end{itemize}
%  \item why we do it: 
%    \begin{itemize}
%    \item getting things right: labelled systems for classical modal
%      logics, or plain propositional intuiionistic logic have one
%      binary relation, which is exactly the relation of the standard
%      Kripke semantics. The standard Kripke semantics for
%      intuitionistic modal logic is birelational, with a strong
%      connectiion between the two binary relations. A clean labelled
%      system for these logics should incorporate both relations.
%    \item We get new logics.  In the standard nested sequent
%      or labelled sequent system for intuitionistic modal logic we
%      always have both conjuncts we we add Scott-Lemmon/Geach axioms;
%      but here we can have them separtely.
%    \end{itemize}
%  \item how we do it:
%    \begin{itemize}
%    \item We start from the Maehara nested sequent system (which can
%      be translated straightforwardly into a labelled system with one
%      relation $R$.
%    \item the rules for $\AND$ and $\OR$ are trivial and as expected
%      they can be presented such that they are invertible on the left
%      and the right, they do not concern the relations $R$ and $\le$. 
%    \item So should be the rules for $\IMP$-left and $\BOX$-left and
%      identity, but there is a subtlety; see below...
%    \item the rules for $\DIA$ are also straightforward, as they
%      concern only $R$ but not $\le$.
%    \item The rules $\BOX$-right and $\IMP$-right are the ones that
%      force single conclusion in the Maehara system. This means that
%      here they create a future, using the relation $\le$.
%    \item In order to make everything work together, this forces us to
%      have reflexivity and transitivity of $\le$, explicely as rules
%      in the system, and have rules for the F1 and F2 conditions
%      making the relations $R$ and $\le$ go together.
%    \item As a consequence we also need a rule realizing monotonicity
%      for the formulas on the right.
%    \item However, we designed the system such that monotonicity is
%      admisible by encorporating it into the rules for identity,
%      $\IMP$-left and $\BOX$-left.
%    \end{itemize}
%  \end{itemize}
%}
%%
%%For sequent systems for intuitionistic logics there is always a choice
%%to be made: make the system \emph{single conclusion} following
%%Gentzen~\cite{Gentzen} or \emph{multiple conclusion} following
%%Maehara~\cite{Maehara}. In our work we choose the multiple conclusion
%%variant because of the closer correspondence to the semantics. In that
%%respect, our system is closer to~\cite{kuz:str} than
%%to~\cite{Simpson} and~\cite{mar:str}.

### 2019-10-03 caleidoscope/main.tex
correcting the soundness file
small changes and comments everywhere

\input{macros2}

@@

 \begin{abstract}
   In this paper we present a labelled sequent system for
   intuitionistic modal logics such that there is not only one, but
   two relation symbols for the labels: one for the accessibility
   relation associated with the Kripke semantics for normal modal logics and
   one for the preorder relation associated with the Kripke semantics
   for intuitionistic logic.
   %%
   This puts our system in close correspondence with the standard
   birelational Kripke semantics for intuitionistic modal logics. As a
   consequence it can encompass a wider range of intuitionistic modal
   logics than standard labelled systems.
   %%
   We also show an internal cut elimination proof for our system.
 \end{abstract}

>>>>

\input{new-macros}

@@

 \begin{abstract}
   In this paper we present a labelled sequent system for
   intuitionistic modal logics such that there is not only one, but
   two relation symbols appearing in sequents: one for the accessibility
   relation associated with the Kripke semantics for normal modal logics and
   one for the preorder relation associated with the Kripke semantics
   for intuitionistic logic.
   %%
   This puts our system in close correspondence with the standard
   birelational Kripke semantics for intuitionistic modal logics. As a
   consequence it can encompass a wider range of intuitionistic modal
   logics than existing labelled systems.
   %%
   We also show an internal cut elimination proof for our system.
 \end{abstract}

### 2019-10-03 caleidoscope/old-text.tex
small changes and comments everywhere

The proof is a careful adaptation of standard techniques.
%
%In particular, in order to prove (i) $\rightarrow$ (ii), we present a syntactic completeness proof with respect the Hilbert system which means that we prove all Hilbert axioms using the rules from our system $\labIKp$: we give a proof for all the axioms of propositional intuionistic logic, for the five variants of $\mathsf{k}$ axiom from the intuitionistic syntax and finally, we simulate the necesssitation rule and modus ponens. As we mentioned, in the course of the proof of (i) $\rightarrow$ (ii), we have to derive the five $\kax$ axioms. 
%
%As an example, we display the derivation of $\kax[4]$ which also illustrates the need of having the rule corresponding to $\rn{F_1}$ in the system.
\bigskip

>>>>

The proof is a careful adaptation of standard techniques.
%
In particular, in order to prove (i) $\rightarrow$ (ii), we present a syntactic completeness proof with respect the Hilbert system which means that we prove all Hilbert axioms using the rules from our system $\labIKp$: we give a proof for all the axioms of propositional intuionistic logic, for the five variants of $\mathsf{k}$ axiom from the intuitionistic syntax and finally, we simulate the necesssitation rule and modus ponens. As we mentioned, in the course of the proof of (i) $\rightarrow$ (ii), we have to derive the five $\kax$ axioms. 
%
As an example, we display the derivation of $\kax[4]$ which also illustrates the need of having the rule corresponding to $\rn{F_1}$ in the system.
\bigskip

### 2019-10-03 caleidoscope/preliminaries.tex
small changes and comments everywhere

%The axiomatisation that is now generally accepted as intuitionistic modal logic $\IK$ was given by Plotkin and Stirling~\cite{Plotkin} and is equivalent to the one proposed by Fischer-Servi~\cite{Fischer}. It is obtained from intuitionistic propositional logic $\mathsf{IPL}$ by adding:
\begin{itemize}
	\item the \emph{necessitation rule}: if $\fm A$ is a theorem then $\fm{\BOX A}$ is also a theorem; and
	\item the following five variants of the \emph{distributivity axiom} $\rn{k}$:
	\begin{equation*}
	\label{eq:ik}%\hskip-2em
	\begin{array}[t]{r@{\;}l@{\qquad}r@{\;}l@{\qquad}r@{\;}l}
	\kax[1]\colon&\fm{\BOX(A\IMP B)\IMP(\BOX A\IMP\BOX B)}
	&
	\kax[3]\colon&\fm{\DIA(A\OR B)\IMP(\DIA A\OR\DIA B)}
	&
	\kax[5]\colon&\fm{\DIA\BOT\IMP\BOT}
	\\
	\kax[2]\colon&\fm{\BOX(A\IMP B)\IMP(\DIA A\IMP\DIA B)}
	&
	\kax[4]\colon&\fm{(\DIA A\IMP \BOX B)\IMP\BOX(A\IMP B)}\\%x[1ex]
	\end{array}
	\end{equation*}
\end{itemize}

@@

\todo{Add something else about Fischer-Servi theorem?}\lutz{no, but we should add the Plotkin-Stirling theorem about the relation between axioms and frame conditions} \todo{add an intro to the next theorem}

%Plotkin and Stirling give proper correspondence results, but they must distinguish the two dual parts of each axiom and appeal to both relations of the frames,

\lutz{maybe the foloowing goes to the ``extensions'' section. otherwise we have to define $\agklmn$ here}
\begin{theorem}[\cite{plotkin:stirling:86}]
	An intuitionistic modal frame $\langle W, \rel, \le \rangle$ validates $\agklmn$ if and only if the frame satisfies:

	if $\lb w R^{k} \lb u$ and $\lb w R^{m} \lb v$ then there exists $\lb{u'}$ such that $\futs{u}{u'}$ and there exists $\lb x$ such that $\lb{u'}R^{l}\lb x$ and $\lb v R^{n} \lb x$.

\end{theorem}

>>>>

%The axiomatisation that is now generally accepted as intuitionistic modal logic $\IK$ was given by Plotkin and Stirling~\cite{Plotkin} and is equivalent to the one proposed by Fischer-Servi~\cite{Fischer}. It is obtained from intuitionistic propositional logic $\mathsf{IPL}$ by adding:
\begin{itemize}
	\item the \emph{necessitation rule}: if $\fm A$ is a theorem then $\fm{\BOX A}$ is also a theorem; and
	\item the following five variants of the \emph{distributivity axiom} $\rn{k}$:
	\begin{equation*}
	\label{eq:ik}%\hskip-2em
	\begin{array}[t]{r@{\;}l@{\qquad}r@{\;}l@{\qquad}r@{\;}l}
	\kax[1]\colon&\fm{\BOX(A\IMP B)\IMP(\BOX A\IMP\BOX B)}
	&
	\kax[3]\colon&\fm{\DIA(A\OR B)\IMP(\DIA A\OR\DIA B)}
	&
	\kax[5]\colon&\fm{\DIA\BOT\IMP\BOT}
	\\
	\kax[2]\colon&\fm{\BOX(A\IMP B)\IMP(\DIA A\IMP\DIA B)}
	&
	\kax[4]\colon&\fm{(\DIA A\IMP \BOX B)\IMP\BOX(A\IMP B)}\\%x[1ex]
	\end{array}
	\end{equation*}
\end{itemize}
\sonia{axioms vs. axiom schemes? do we need to write these on atomic formulas $a,b$ instead of generic formulas $A,B$?}

@@

### 2019-10-03 caleidoscope/system.tex
small changes and comments everywhere

In this section we present our fully labelled sequent proof system
for intuitionistic modal logics. The starting point is the notion of a
\emph{labelled formula} which is a pair $\labels xA$ of a label $\lb
x$ and a formula $\fm A$. The intended meaning is that $\fm A$ holds
at world $\lb x$. A \emph{relation atom} is either an expression
$\accs xy$ or $\futs xy$ where $\lb x$ and $\lb y$ are labels and the
intended meaning is that the corresponing relation holds between the
worlds $\lb x$ and $\lb y$. This is enough to define the notion of
\emph{(labelled) sequent}, which is a triple $\lseq\B\Left\Right$,
where $\B$ is a set of relational and pre-order atoms, and $\Left$ and $\Right$ are
multi-sets of labelled formulas, all written as list, separated by commas.

@@

Now we can present the inference rules for \emph{system $\labIKp$} for the logic $\IK$.
We obtained this system, shown in Figure~\ref{fig:labIKp}, as follows.
Our starting point was the multiple-conlusion nested sequent system \emph{\`a la} Maehara (as presented in~\cite{str:2017maehara}), which can straightforwardly translated into the labelled setting, and which yields the rules $\llabrn\bot$, $\llabrn\AND$, $\rlabrn\AND$, $\llabrn\OR$, $\rlabrn\OR$, $\llabrn\DIA$, and $\rlabrn\DIA$ as we show them in Figure~\ref{fig:labIKp}. However, this naive translation wpould also yield the rules $\rn{id'}$, $\llabrn\IMP'$, and~$\llabrn\BOX'$:
\begin{equation}
  \label{eq:nomon}
  \vlinf{\rn{id}'}{}{\B, \Left, \labels{x}{a} \SEQ \Right, \labels{x}{a} }{}
  \quad
  \vliinf{\llabrn\IMP'}{}{\B,\Left, \labels{x}{A \IMP B} \SEQ \Right}{\B,\labels{x}{A \IMP B}, \Left \SEQ \Right, \labels{x}{A}}{\B, \Left, \labels{x}{B} \SEQ \Right}
  \quad
  \vlinf{\llabrn\BOX}{}{\B, \accs xz, \Left, \labels{x}{\BOX A} \SEQ \Right}{\B, \accs xz, \Left, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right}
\end{equation}
that are not sufficient for a complete system, and we will see below
why. Before, let us first look at the rules $\rlabrn\IMP$ and
$\rlabrn\BOX$. In the multiple-conlusion nested sequent system
of~\cite{str:2017maehara}, these are the two rules that force
single-conclusion. This is the reason why in our system we need the
$\le$ in the relational atoms; we need to go to the future of the
current world.  In the Kripke-semantics in~\eqref{eq:kripke} the two
connectives $\IMP$ and $\BOX$ are the ones that need the
future-relation $\le$. This relation is reflexive and transitive. In
order to capture that in the proof system, we need to add the rules
$\rn{refl}$ and $\rn{trans}$.

@@

One can prove this proposition in the same way as one usually proves
admissibity of contraction in a sequent calculus, by induction on the
height of the derivation, which in fact would yield a stronger result,
namely that $\llabrn{mon}$ and $\rlabrn{mon}$ are \emph{height
  preserving} admissible for $\labIKp$. However, we do not need this
result in this paper, and therefore we leave it as an exercise to the
reader. Nonetheless, we will give a short proof of
Proposition~\ref{prop:mon-adm} at the end of this section.

>>>>

In this section we present our fully labelled sequent proof system
for intuitionistic modal logics. The starting point is the notion of a
\emph{labelled formula} which is a pair $\labels xA$ of a label $\lb
x$ and a formula $\fm A$. The intended meaning is that $\fm A$ holds
at world $\lb x$. A \emph{relation atom} is either an expression
$\accs xy$ or $\futs xy$ where $\lb x$ and $\lb y$ are labels and the
intended meaning is that the corresponing relation holds between the
worlds $\lb x$ and $\lb y$. 
\sonia{Not sure we want to make it sound so semantical here, with not much context.}
%
This is enough to define the notion of
\emph{(labelled) sequent}, which is a triple $\lseq\B\Left\Right$,
where $\B$ is a set of relational and pre-order atoms, 
\sonia{inconsistent naming with intro}
and $\Left$ and $\Right$ are
multi-sets of labelled formulas, all written as list, separated by commas.

@@

Now we can present the inference rules for \emph{system $\labIKp$} for the logic $\IK$.
We obtained this system, shown in Figure~\ref{fig:labIKp}, as follows.
\sonia{This is an interesting alternative view but for me the starting point really is the Maffezioli, Naibo and Negri paper. Maybe rephrase slightly to acknowledge their work too.}
Our starting point was the multiple-conlusion nested sequent system \emph{\`a la} Maehara (as presented in~\cite{str:2017maehara}), which can straightforwardly translated into the labelled setting, and which yields the rules $\llabrn\bot$, $\llabrn\AND$, $\rlabrn\AND$, $\llabrn\OR$, $\rlabrn\OR$, $\llabrn\DIA$, and $\rlabrn\DIA$ as we show them in Figure~\ref{fig:labIKp}. However, this naive translation wpould also yield the rules $\rn{id'}$, $\llabrn\IMP'$, and~$\llabrn\BOX'$:
\begin{equation}
  \label{eq:nomon}
  \vlinf{\rn{id}'}{}{\B, \Left, \labels{x}{a} \SEQ \Right, \labels{x}{a} }{}
  \quad
  \vliinf{\llabrn\IMP'}{}{\B,\Left, \labels{x}{A \IMP B} \SEQ \Right}{\B,\labels{x}{A \IMP B}, \Left \SEQ \Right, \labels{x}{A}}{\B, \Left, \labels{x}{B} \SEQ \Right}
  \quad
  \vlinf{\llabrn\BOX}{}{\B, \accs xz, \Left, \labels{x}{\BOX A} \SEQ \Right}{\B, \accs xz, \Left, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right}
\end{equation}
that are not sufficient for a complete system, and we will see below
why. Before, let us first look at the rules $\rlabrn\IMP$ and
$\rlabrn\BOX$. In the multiple-conlusion nested sequent system
of~\cite{str:2017maehara}, these are the two rules that force
single-conclusion. This is the reason why in our system we need the
$\le$ in the relational atoms; we need to go to the future of the
current world.  In the Kripke-semantics in~\eqref{eq:kripke} the two
connectives $\IMP$ and $\BOX$ are the ones that need the
future-relation $\le$. This relation is reflexive and transitive. In
order to capture that in the proof system, we need to add the rules
$\rn{refl}$ and $\rn{trans}$.

@@

One can prove this proposition in the same way as one usually proves
admissibity of contraction in a sequent calculus, by induction on the
height of the derivation, which in fact would yield a stronger result,
namely that $\llabrn{mon}$ and $\rlabrn{mon}$ are \emph{height
  preserving} admissible for $\labIKp$. However, we do not need this
result in this paper, and therefore we leave it to the interested
reader. Nonetheless, we will give a short proof of
Proposition~\ref{prop:mon-adm} at the end of this section.
